---
title: "Research Analysis for NOAA Weather"
output:
  html_notebook: default
  html_document: default
  md_document:
    variant: markdown_github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Storm Research Analysis {.tabset .tabset-fade .tabset-pills}

### Main Report Section

##### Synopsis 
This report utilizes NOAA storm weather data to explain weather patterns. The weather in research are disastrous conditions, and the attributes of the weather. The analysis will also deliver results on the events that directly may have resulted from the unusual weather.

##### Data Processing

The data was fetched from NOAA (see Citation for url). The source to pull process and retain the data can be see in the code below.

```{r Data Processing, include=TRUE, echo=TRUE}
### Week 1 Project ###
library(data.table)
webURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destURL <- file.path(getwd(),paste("repdata_data_StormData.csv",".bz2",sep = ""))
download.file(webURL,destfile = destURL)

#unzip(destURL)
#localFileURL <- file.path(getwd(),paste("repdata_data_StormData",".csv",sep = ""))
repdata <- read.csv(destURL)
```

##### Summary

You can also embed plots, for example:

```{r pressure, include = TRUE, echo=FALSE}
as.table(summary(repdata))

## rmarkdown::paged_table()
```

### Source Citation and Resource:
    NATIONAL WEATHER SERVICE INSTRUCTION  10-1605 
    AUGUST 17, 2007 
    Operations and Services 
    Performance, NWSPD 10-16
    STORM DATA PREPARATION


    Storm Data Documentation: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
    FAQ: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
    

    
### License and Copyright Notice
    ### Default License and Copyright Notice
    Self published Research Analysis for NOAA Weather
    Copyright (C) 2017 Michael Garcia. All Rights Reserved.
    Unless otherwise noted, default copyright is excercised
    and exclusive rights of the author. 
    https://www.copyright.gov/title17/
    
    Source code from other authors, and only that source code,
    retain their respective License. If any conflict overall;
    the most permissive, copy-left license overrides project License.
    Fees and permits by author, if any, still apply.
    
    Licensing, pricing, permission inquiries, token to use source code 
      contact mgar_datascience at protonmail.com.
    
    To submit contributions required by permissive copy-left license,
      contact mgar_datascience at protonmail.com for token to 
      source repository.

    This program is distributed  WITHOUT ANY WARRANTY; without even the 
    implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
    PURPOSE.  
    
##
