---
title: "Research Analysis for NOAA Weather"
output:
  md_document:
    variant: markdown_github
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
### Week 1 Project ###
library(data.table)


webURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destURL <- file.path(getwd(),paste("repdata_data_StormData.csv",".bz2",sep = ""))
download.file(webURL,destfile = destURL)

#unzip(destURL)
#localFileURL <- file.path(getwd(),paste("repdata_data_StormData",".csv",sep = ""))
repdata <- read.csv(destURL)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



## Source Citation and Resource:
    NATIONAL WEATHER SERVICE INSTRUCTION  10-1605 
    AUGUST 17, 2007 
    Operations and Services 
    Performance, NWSPD 10-16
    STORM DATA PREPARATION


Storm Data Documentation: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
FAQ: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
