---
title: "Research Analysis for NOAA Weather"
output:
  md_document:
    variant: markdown_github
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Storm Research Analysis {.tabset .tabset-fade .tabset-pills}

### Main Report Section

##### Synopsis 
This report utilizes NOAA storm weather data to explain weather patterns. The weather in research are disastrous conditions, and the attributes of the weather. The analysis will also deliver results on the events that directly may have resulted from the unusual weather.

##### Data Processing

The data was fetched from NOAA (see Citation for url). The source to pull process and retain the data can be see in the code below.

```{r Data Processing, include=TRUE, echo=TRUE}
### Week 1 Project ###
library(data.table)
webURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destURL <- file.path(getwd(),paste("repdata_data_StormData.csv",".bz2",sep = ""))
download.file(webURL,destfile = destURL)

#unzip(destURL)
#localFileURL <- file.path(getwd(),paste("repdata_data_StormData",".csv",sep = ""))
repdata <- read.csv(destURL)
```

##### Summary

You can also embed plots, for example:

```{r pressure, include = TRUE, echo=FALSE}
as.table(summary(repdata))
```

### Source Citation and Resource:
    NATIONAL WEATHER SERVICE INSTRUCTION  10-1605 
    AUGUST 17, 2007 
    Operations and Services 
    Performance, NWSPD 10-16
    STORM DATA PREPARATION


    Storm Data Documentation: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
    FAQ: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
    

    
### License and Copyright Notice
    ### License and Copyright Notice
    Research Analysis for NOAA Weather
    Copyright (C) 2017 Michael Garcia. All Rights Reserved.
    
    Licensing Inquiries and Requests: mgar_datascience at protonmail.com

    This program is free software: you can redistribute it and/or modify
    it under the terms of the GNU Affero General Public License as published
    by the Free Software Foundation, either version 3 of the License, or
    (at your option) any later version.

    This program is distributed in the hope that it will be useful,
    but WITHOUT ANY WARRANTY; without even the implied warranty of
    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
    GNU Affero General Public License for more details.

    You should have received a copy of the GNU Affero General Public License
    along with this program.  If not, see <http://www.gnu.org/licenses/>.
    
##
